---
title: "Data Proposal"
author: "Andrew Graham, Ji Hoon Justin Han, Ethan Nash"
date: '2022-12-13'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# R Packages 
This list includes all of the packages we used for our analyses.
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(MASS)
library(effects)
library(hrbrthemes)
library(ggpubr)
library(lme4)
library(car)
library(broom)
library(rstatix)
library(ggfortify)
library(MuMIn)
```
# Tidying and Cleaning Our Data
1: Individual csv files separated by species were uploaded into R. Species files were named as the first two letters of Genus and Species (ex: ACBL = **Ac**anthurus **bl**ochii).
```{r}
ACBL <- read.csv("ACBL.csv")
ACNF <- read.csv("ACNF.csv")
ACTR <- read.csv("ACTR.csv")
CAME <- read.csv("CAME.csv")
NALI <- read.csv("NALI.csv")
SCPS <- read.csv("SCPS.csv")
SCRU <- read.csv("SCRU.csv")
THDU <- read.csv("THDU.csv")
```
2. Raw files were combined into one dataset
```{r}
combined_raw <- bind_rows(ACBL,ACNF, ACTR, CAME, NALI, SCPS, SCRU, THDU)
```
3. combined raw files renamed
```{r}
fish_data <- combined_raw
```
4. Simplified and renamed column
```{r}
colnames (fish_data) [1] <- "Observer"
colnames(fish_data) <- c("Observer","Code", "Family", "Genus & Species", "Trophic_level", "Fish_size", "School_size", "Start_distance", "FID", "Activity", "Response", "MAD", "Site", "Covid")
```
5. Unnecessary columns removed
```{r}
fish_data <- subset(fish_data, select=-c(Code, Observer))
```
6. Separated "Genus & Species" column into two separate columns
```{r}
fish_data <- fish_data %>% separate(`Genus & Species`, c('Genus', 'Species'))
```
7. Added column of log-transformed
```{r}
fish_data$logFID <- log(fish_data$FID)
```
8. Some character data replaced for clarification
```{r}
fish_data <- fish_data %>%
  mutate(Covid = str_replace(Covid, "Pre", "Closed"))
fish_data <- fish_data %>%
  mutate(Covid = str_replace(Covid, "Post", "Open"))
```
9. Blanks in dataset were replaced with N/A
```{r}
fish_data  <- replace(fish_data , fish_data =='', NA)
```
10. Character columns changed to factors
```{r}
fish_data$Covid <- as.factor(fish_data$Covid)
fish_data$Trophic_level <- as.factor(fish_data$Trophic_level)
fish_data$Activity <- as.factor(fish_data$Activity)
fish_data$Response <- as.factor(fish_data$Response)
fish_data$Site <- as.factor(fish_data$Site)
fish_data$Family <- as.factor(fish_data$Family)
fish_data$Genus <- as.factor(fish_data$Genus)
fish_data$Species <- as.factor(fish_data$Species)
```
10. Final Checks
```{r}
glimpse(fish_data)
```
# Data Visualization
Colorblind Palette colors added in
```{r}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```
Box plot of Flight Initiation Distance by bay access to Hanauma Bay and by species
```{r}
ggplot(fish_data, aes(x = log(FID), Covid)) +
  geom_boxplot(na.rm = TRUE) +
  scale_fill_manual(values=cbPalette) +
  ggtitle("Flight Initiation Distance by Access") +
  xlab("FID (mm)") +
  ylab("Bay Status")
ggplot(fish_data, aes(x = FID, Species, fill = Covid)) + 
  geom_boxplot() +
  scale_fill_manual(values=cbPalette) +
  ggtitle("Flight Initiation Distance x spp. by Bay Access") +
  xlab("FID (mm)")
```
Box plot of Minimum Approach Distance by bay access to Hanauma Bay by species
```{r}
ggplot(fish_data, aes(x = log(MAD), Covid)) +
  geom_boxplot(na.rm = TRUE) +
  scale_fill_manual(values=cbPalette) +
  ggtitle("Minimal Approach Distance by Access") +
  xlab("MAD (mm)") +
  ylab("Bay Status")
ggplot(fish_data, aes(x = MAD, y = Species, fill = Covid)) + 
  geom_boxplot(na.rm = TRUE) +
  scale_fill_manual(values=cbPalette) +
  ggtitle("Minimal Approach Distance x spp. by Bay Access") +
  xlab("MAD (mm)")
```
Box plot of Flight Initiation Distance & Minimal Approach Distance by Trophic level
```{r}
ggplot(fish_data, aes(x = FID, Y=Trophic_level)) + 
  geom_boxplot() +
  ggtitle("Flight Initiation Distance x Trophic level") +
  xlab("FID (mm)")
ggplot(fish_data, aes(x = MAD, Y=Trophic_level)) + 
  geom_boxplot(na.rm = TRUE) +
  ggtitle("Minimal Approach Distance x Trophic level") +
  xlab("MAD (mm)")
```
Scatterplot showing Fish Size and FID
```{r}
ggplot(data = fish_data, mapping = aes(x=FID, y= Fish_size)) +
  geom_point(size = 1, na.rm = TRUE) +
  geom_smooth(formula = y ~ x, method ="lm", color = "red", se=FALSE, na.rm = TRUE) +
  stat_regline_equation(label.y = 1000, aes(label = ..rr.label..)) +
  facet_wrap( ~ Species, ncol=2) +
  ggtitle("Fish size vs. FID") +
  xlab("Fish length (mm)") +
  ylab("FID (mm)")
```
Scatterplot showing MAD by Fish Size
```{r}
ggplot(data = fish_data, mapping = aes(x=MAD, y= Fish_size)) +
  geom_point(size = 1, na.rm = TRUE) +
  geom_smooth(formula = y ~ x, method ="lm", color = "blue", se=FALSE, na.rm = TRUE) +
  stat_regline_equation(label.y = 1500, aes(label = ..rr.label..)) +
  facet_wrap( ~ Species, ncol=2) +
  ggtitle("Minimal Approach Distance by Fish size") +
  xlab("Fish length (mm)") +
  ylab("MAD (mm)")
```


