---
title: "Final Project"
author: "Andrew Graham, Ji Hoon Justin Han, Ethan Nash"
date: '2022-12-13'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Packages 
This list includes all of the packages we used for our analyses.
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(MASS)
library(effects)
library(hrbrthemes)
library(ggpubr)
library(lme4)
library(car)
library(broom)
library(rstatix)
library(ggfortify)
library(MuMIn)
library(moderndive)
library(lmtest)
library(zoo)
```
# Tidying and Cleaning Our Data
1: Individual csv files separated by species were uploaded into R. Species files were named as the first two letters of Genus and Species (ex: ACBL = **Ac**anthurus **bl**ochii).
```{r}
ACBL <- read.csv("ACBL.csv")
ACNF <- read.csv("ACNF.csv")
ACTR <- read.csv("ACTR.csv")
CAME <- read.csv("CAME.csv")
NALI <- read.csv("NALI.csv")
SCPS <- read.csv("SCPS.csv")
SCRU <- read.csv("SCRU.csv")
THDU <- read.csv("THDU.csv")
```
2. Raw files were combined into one dataset
```{r}
combined_raw <- bind_rows(ACBL,ACNF, ACTR, CAME, NALI, SCPS, SCRU, THDU)
```
3. combined raw files renamed
```{r}
fish_data <- combined_raw
```
4. Simplified and renamed column
```{r}
colnames (fish_data) [1] <- "Observer"
colnames(fish_data) <- c("Observer","Code", "Family", "Genus & Species", "Trophic_level", "Fish_size", "School_size", "Start_distance", "FID", "Activity", "Response", "MAD", "Site", "Covid")
```
5. Unnecessary columns removed
```{r}
fish_data <- subset(fish_data, select=-c(Code, Observer))
```
6. Separated "Genus & Species" column into two separate columns
```{r}
fish_data <- fish_data %>% separate(`Genus & Species`, c('Genus', 'Species'))
```
7. Added column of log-transformed
```{r}
fish_data$logFID <- log(fish_data$FID)
```
8. Some character data replaced for clarification
```{r}
fish_data <- fish_data %>%
  mutate(Covid = str_replace(Covid, "Pre", "Closed"))
fish_data <- fish_data %>%
  mutate(Covid = str_replace(Covid, "Post", "Open"))
```
9. Blanks in dataset were replaced with N/A
```{r}
fish_data  <- replace(fish_data , fish_data =='', NA)
```
10. Character columns changed to factors
```{r}
fish_data$Covid <- as.factor(fish_data$Covid)
fish_data$Trophic_level <- as.factor(fish_data$Trophic_level)
fish_data$Activity <- as.factor(fish_data$Activity)
fish_data$Response <- as.factor(fish_data$Response)
fish_data$Site <- as.factor(fish_data$Site)
fish_data$Family <- as.factor(fish_data$Family)
fish_data$Genus <- as.factor(fish_data$Genus)
fish_data$Species <- as.factor(fish_data$Species)
```
10. Final Checks
```{r}
glimpse(fish_data)
```
# Data Visualization
Colorblind Palette colors added in
```{r}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```
Box plot of Flight Initiation Distance by bay access to Hanauma Bay and by species
```{r}
ggplot(fish_data, aes(x = log(FID), Covid)) +
  geom_boxplot(na.rm = TRUE) +
  scale_fill_manual(values=cbPalette) +
  ggtitle("Flight Initiation Distance by Access") +
  xlab("FID (mm)") +
  ylab("Bay Status")
ggplot(fish_data, aes(x = FID, Species, fill = Covid)) + 
  geom_boxplot() +
  scale_fill_manual(values=cbPalette) +
  ggtitle("Flight Initiation Distance x spp. by Bay Access") +
  xlab("FID (mm)")
```
Box plot of Minimum Approach Distance by bay access to Hanauma Bay by species
```{r}
ggplot(fish_data, aes(x = log(MAD), Covid)) +
  geom_boxplot(na.rm = TRUE) +
  scale_fill_manual(values=cbPalette) +
  ggtitle("Minimal Approach Distance by Access") +
  xlab("MAD (mm)") +
  ylab("Bay Status")
ggplot(fish_data, aes(x = MAD, y = Species, fill = Covid)) + 
  geom_boxplot(na.rm = TRUE) +
  scale_fill_manual(values=cbPalette) +
  ggtitle("Minimal Approach Distance x spp. by Bay Access") +
  xlab("MAD (mm)")
```
Box plot of Flight Initiation Distance & Minimal Approach Distance by Trophic level
```{r}
ggplot(fish_data, aes(x = FID, Y=Trophic_level)) + 
  geom_boxplot() +
  ggtitle("Flight Initiation Distance x Trophic level") +
  xlab("FID (mm)")
ggplot(fish_data, aes(x = MAD, Y=Trophic_level)) + 
  geom_boxplot(na.rm = TRUE) +
  ggtitle("Minimal Approach Distance x Trophic level") +
  xlab("MAD (mm)")
```
Scatterplot showing Fish Size and FID
```{r}
ggplot(data = fish_data, mapping = aes(x=FID, y= Fish_size)) +
  geom_point(size = 1, na.rm = TRUE) +
  geom_smooth(formula = y ~ x, method ="lm", color = "red", se=FALSE, na.rm = TRUE) +
  stat_regline_equation(label.y = 1000, aes(label = ..rr.label..)) +
  facet_wrap( ~ Species, ncol=2) +
  ggtitle("Fish size vs. FID") +
  xlab("Fish length (mm)") +
  ylab("FID (mm)")
```
Scatterplot showing MAD by Fish Size
```{r}
ggplot(data = fish_data, mapping = aes(x=MAD, y= Fish_size)) +
  geom_point(size = 1, na.rm = TRUE) +
  geom_smooth(formula = y ~ x, method ="lm", color = "blue", se=FALSE, na.rm = TRUE) +
  stat_regline_equation(label.y = 1500, aes(label = ..rr.label..)) +
  facet_wrap( ~ Species, ncol=2) +
  ggtitle("Minimal Approach Distance by Fish size") +
  xlab("Fish length (mm)") +
  ylab("MAD (mm)")
```
###Analysis###
Q1. Did fish behavior different by the bay accessibility?

First we checked the normality of the variable of interest (FID) 
```{r}
hist(fish_data$FID) # data right skewed
hist(fish_data$logFID) #log transformation
ggplot(fish_data, aes(x = logFID)) +
  geom_histogram(bins=35, aes(y=..density..), colour="black", fill="white")+
  geom_density(alpha=.2, fill="blue") +
  ggtitle("Histogram of Fligh Initiation Distance") +
  xlab("logFID (mm)")
```
normality check
```{r}
shapiro.test(fish_data$logFID) #normality check failed
ggplot(fish_data, aes(sample = fish_data$logFID)) + 
  stat_qq() +
  stat_qq_line(col = "red")
```
Box plot of FID by Access
```{r}
ggplot(fish_data, aes(x = Covid, y =FID)) +
  geom_boxplot()+
  ggtitle("Flight Initiation Distance by Access") +
  xlab("Bay Access") +
  ylab("FID (mm)")
```
Two-sample t-test of Flight Initiation Distance by Bay Accessibility
```{r}
print(t.test(logFID ~ Covid, data = fish_data))
```

```{r}
covid_lm <- lm(FID~Covid + 0, fish_data)
covid_lm$coefficients
```

Investigated FID by Bay Access by species
```{r}
ggplot(fish_data, aes(x = Covid, y = logFID, fill = Covid)) +
  geom_boxplot()+
  facet_wrap( ~ Species, ncol=2) +
  scale_fill_manual(values=cbPalette) +
  ggtitle("Flight Initiation Distance by Access") +
  xlab("Bay Access") +
  ylab("FID (mm)")
```
```{r}
T.test_ABCL <- fish_data %>%
  filter(Species == "blochii") %>% 
  dplyr::select(MAD, Covid) %>%
  t.test(MAD ~ Covid, data = .)
T.test_ACTR <- fish_data %>%
  filter(Species == "triostegus") %>% 
  dplyr::select(MAD, Covid) %>%
  t.test(MAD ~ Covid, data = .)
T.test_ACNF <- fish_data %>%
  filter(Species == "nigrofuscus") %>% 
  dplyr::select(MAD, Covid) %>%
  t.test(MAD ~ Covid, data = .)
T.test_CAME <- fish_data %>%
  filter(Species == "melampygus") %>% 
  dplyr::select(MAD, Covid) %>%
  t.test(MAD ~ Covid, data = .)
T.test_NALI <- fish_data %>%
  filter(Species == "lituratus") %>% 
  dplyr::select(MAD, Covid) %>%
  t.test(MAD ~ Covid, data = .)
T.test_THDU <- fish_data %>%
  filter(Species == "duperrey") %>% 
  dplyr::select(MAD, Covid) %>%
  t.test(MAD ~ Covid, data = .)
T.test_SCPS <- fish_data %>%
  filter(Species == "psittacus") %>% 
  dplyr::select(MAD, Covid) %>%
  t.test(MAD ~ Covid, data = .)
T.test_SCRU <- fish_data %>%
  filter(Species == "rubroviolaceus") %>% 
  dplyr::select(MAD, Covid) %>%
  t.test(MAD ~ Covid, data = .)

t_spp <-c("blochii","triostegus","nigrofuscus","melampygus","lituratus","duperrey", "psittacus", "rubroviolaceus")
t_pval <-c(T.test_ABCL$p.value,T.test_ACTR$p.value, T.test_NALI$p.value, T.test_CAME$p.value, T.test_ACNF$p.value, T.test_ABCL$p.value, T.test_SCPS$p.value, T.test_SCRU$p.value)
data.frame(t_spp, t_pval)
```
Q2. Did the bay closure effect FID by site (sector of the bay)?
```{r}
ggplot(fish_data, aes(x = Site, y = logFID, fill = Covid)) +
  geom_boxplot()+
  ggtitle("Flight Initiation Distance by Site") +
  xlab("Site") +
  ylab("FID (mm)")
```
```{r}
t_site_backdoor <- fish_data %>%
  filter(Site == "Backdoor") %>% 
  dplyr::select(logFID, Covid) %>%
  t.test(logFID ~ Covid, data = .)
t_site_Chennel <- fish_data %>%
  filter(Site == "Channel") %>% 
  dplyr::select(logFID, Covid) %>%
  t.test(logFID ~ Covid, data = .)
t_site_Keyhole <- fish_data %>%
  filter(Site == "Keyhole") %>% 
  dplyr::select(logFID, Covid) %>%
  t.test(logFID ~ Covid, data = .)
t_site_WitchesBrew <-fish_data %>%
  filter(Site == "Witches_Brew") %>% 
  dplyr::select(logFID, Covid) %>%
  t.test(logFID ~ Covid, data = .)

t_site <-c("Backdoor","Channel","Keyhole","WitchesBrew" )
t_site_pval <-c(t_site_backdoor$p.value,t_site_Chennel$p.value, t_site_Keyhole$p.value,t_site_WitchesBrew$p.value)
data.frame(t_site, t_site_pval)
```
Q3. Does fish behavior differ by site (sector)?

```{r}
ggplot(fish_data, aes(x = Site, y = logFID)) +
  geom_boxplot()+
  ggtitle("Flight Initiation Distance by Site") +
  xlab("Site") +
  ylab("FID (mm)")
```
FID vary by species?
```{r}
ggplot(fish_data, aes(Species, FID, fill = Species)) + 
  geom_boxplot() + 
  stat_summary(fun = mean)

FID_Site_aov <- aov(FID ~ Site, data = fish_data)
summary(FID_Site_aov)

TukeyHSD(FID_Site_aov) #post hoc

```

MAD vary by species?
```{r}
ggplot(fish_data, aes(Species, MAD, fill = Species)) + 
  geom_boxplot() + 
  stat_summary(fun = mean)

MAD_Site_aov <- aov(MAD ~ Site, data = fish_data)
summary(MAD_Site_aov)

TukeyHSD(MAD_Site_aov) #post hoc
```

Q4. Does fish size determine fish behavior?
```{r}
fish_data %>%
  filter(Fish_size < 1000) %>%
  ggplot(aes(FID, Fish_size)) + 
  geom_point() + 
  geom_parallel_slopes(se=FALSE) +
  stat_regline_equation(label.y = 1100, aes(label = ..rr.label..))
```
```{r}
fish_data %>%
  filter(Fish_size < 1000) %>%
  ggplot(aes(FID, Fish_size, color=Species)) + 
  geom_point() + 
  geom_parallel_slopes(se=FALSE)
```
```{r}
fish_data %>%
  filter(Fish_size < 1000) %>%
  ggplot(data = fish_data, mapping = aes(x=FID, y= Fish_size)) +
  geom_point(size = 1, na.rm = TRUE) +
  geom_smooth(formula = y ~ x, method ="lm", color = "red", se=FALSE, na.rm = TRUE) +
  stat_regline_equation(label.y = 1100, aes(label = ..rr.label..)) +
  facet_wrap( ~ Species, ncol=3) +
  ggtitle("Fish size vs. FID") +
  xlab("Fish length (mm)") +
  ylab("FID (mm)")
```
```{r}
fish_data %>%
  filter(Fish_size < 1000) %>%
  ggplot(aes(MAD, Fish_size)) + 
  geom_point() + 
  geom_parallel_slopes(se=FALSE) +
  stat_regline_equation(label.y = 1100, aes(label = ..rr.label..))
fish_data %>%
  filter(Fish_size < 1000) %>%
  ggplot(aes(MAD, Fish_size, color=Species)) + 
  geom_point() + 
  geom_parallel_slopes(se=FALSE)
fish_data %>%
  filter(Fish_size < 1000) %>%
  ggplot(mapping = aes(x=MAD, y= Fish_size)) +
  geom_point(size = 1, na.rm = TRUE) +
  geom_smooth(formula = y ~ x, method ="lm", color = "red", se=FALSE, na.rm = TRUE) +
  stat_regline_equation(label.y = 1000, aes(label = ..rr.label..)) +
  facet_wrap( ~ Species, ncol=3) +
  ggtitle("Fish size vs. MAD") +
  xlab("Fish length (mm)") +
  ylab("MAD (mm)")
```
```{r}
fish_data1 <- na.omit(fish_data)

#linear regression
FID_FISHSIZE <- lm(logFID ~ Fish_size, data = fish_data1)
summary(FID_FISHSIZE)$adj.r.squared
FID_MAD <- lm(logFID ~ MAD, data = fish_data1)
summary(FID_MAD)$adj.r.squared
FID_SITE <- lm(logFID ~ Site, data = fish_data1)
summary(FID_SITE)$adj.r.squared
```
Model selection
```{r}
fish_data1 <- na.omit(fish_data)

mod.1 <- lm(logFID ~ Fish_size + Species + Fish_size:Species + log(MAD) + Site + Start_distance, data = fish_data1)
mod.2 <- lm(logFID ~ Fish_size + Species + Fish_size:Species + log(MAD) + Start_distance, data = fish_data1)
mod.3 <- lm(logFID ~ Fish_size + log(MAD) + Fish_size:Species + Start_distance, data = fish_data1)
mod.4 <- lm(logFID ~ log(MAD) + Start_distance, data = fish_data1)
mod.5 <- lm(logFID ~ log(MAD), data = fish_data1)

model.sel(list(mod.1, mod.2, mod.3, mod.4, mod.5), rank = AIC)

summary(mod.2)
```
Friendly fish???
```{r}
Friendly_lm <- lm(MAD ~ FID + 0, data = fish_data1)
summary(Friendly_lm)

ggplot(data = fish_data, mapping = aes(x=MAD, y= FID)) +
  geom_point(size = 1, na.rm = TRUE) +
  geom_smooth(formula = y ~ x, method ="lm", color = "red", se=FALSE, na.rm = TRUE) +
  stat_regline_equation(label.y = 5000, aes(label = ..rr.label..)) +
  ggtitle("Friendly Fish") +
  xlab("MAD (mm)") +
  ylab("FID (mm)")

ggplot(data = fish_data, mapping = aes(x=MAD, y= FID)) +
  geom_point(size = 1, na.rm = TRUE) +
  geom_smooth(formula = y ~ x, method ="lm", color = "red", se=FALSE, na.rm = TRUE) +
  stat_regline_equation(label.y = 5000, aes(label = ..rr.label..)) +
  facet_wrap( ~ Species, ncol=3)
```